---
title: "ETC3555 2018 - Lab 1"
author: "Cameron Roach"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

## Preliminaries

This lab will focus on working with data. We will cover

* Loading data into `R`
* Reshaping data frames
* Creating new variables
* Fitting models to multiple data sets.

This lab has been designed to reflect a real world scenario where a worker may be required to predict electricity demand for multiple buildings using monthly electricity consumption data.

Before proceeding, ensure you have the latest version of `R` and RStudio installed on your computer. Install the required packages by running

```{r eval=FALSE}
install.packages(c("tidyverse", "randomForest"))
```

## Exercises

### Exercise 1

Download the raw data files from __TODO__ and write a function to read them into `R`. Reshape the data into a tidy format. If you are unfamiliar with tidy datasets, please refer to Hadley Wickham's paper on the subject (available [here](https://www.jstatsoft.org/article/view/v059i10/v59i10.pdf)).

### Exercise 2

Adjust the below code to visualise monthly electricity demand for several of the buildings. Explain what each line of code does. What does the pipe operator, `%>%` do? How does this differ to the `+` method?

_Note: Plotting with the `ggplot2` package will be explored in detail in Lab 2._

```{r eval = FALSE}
elec_df %>% 
  filter(BID %in% sample(BID, 9)) %>%  
  gather(variable, value, kwh, temperature) %>% 
  ggplot(elec_df, aes(x = dt, y = kwh, colour = variable)) +
  geom_line() +
  facet_wrap(~BID, scales = "free_y") +
  labs(title = "Temperature and electricity demand.",
       x = "kWh",
       y = "Date")
```

### Exercise 3

We have now seen a simple example that uses `dplyr` and `tidyr` functions. The `dplyr` package is used for summarising and creating new variables. The `tidyr` package is used for reshaping data frames. Use `dplyr`'s `mutate` function to add hour of day and season variables. Should these new variables be nominal, ordinal or numeric? Can you think of any other predictors that may be useful?

### Exercise 4

Use `purrr`'s `nest` and `map` functions to fit separate random forest models to each building and store each model in a new column of the data frame. _Hint: this should only take a few lines of code._

### Exercise 5

Cross validation using `modelr`.

### Exercise 6

Extract CV RMSE scores and add to new column.