---
title: "ETC3555 2018 - Lab 9 solutions"
subtitle: "Recommender Systems"
author: "Cameron Roach and Souhaib Ben Taieb"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
rm(list=ls())
```

In this assignment, you will implement the collaborative filtering learning algorithm and apply it to a dataset of movie ratings. This dataset consists of ratings on a scale of 1 to 5. The dataset has $n_u = 943$ users, and $n_m = 1682$ movies.

```{r echo = TRUE}
load("movies_ratings.Rda")
list2env(data,.GlobalEnv) # "Y" and "R"
rm(data)

#  Y is a 1682x943 matrix, containing ratings (1-5) of 1682 movies on 
#  943 users
#
#  R is a 1682x943 matrix, where R[i,j] <- 1 if and only if user j gave a
#  rating to movie i
```

The following code gives all movies titles.

```{r echo = TRUE}
source("loadMovieList.R")
movieList <- loadMovieList()
```

(1 mark) Compute the average ratings for "Toy Story (1995)" and "Alaska (1996)".

```{r echo = TRUE}
# ----------------------- YOUR CODE HERE -----------------------
library(tidyverse)

ifelse(R, Y, NA) %>% 
  t() %>% 
  magrittr::set_colnames(movieList) %>% 
  as_data_frame() %>% 
  select("Toy Story (1995)", "Alaska (1996)") %>% 
  summarise_all(funs(mean(., na.rm=TRUE)))
# --------------------------------------------------------------
```

(1 mark) Which user has rated more movies? Plot a histogram of her/his ratings?

```{r echo = TRUE}
# ----------------------- YOUR CODE HERE -----------------------
id_more <- which.max(apply(R, 2, sum))
id_reviewed <- which(R[, id_more] == 1)
print(id_more)
print(length(Y[id_reviewed, id_more]))
hist(Y[id_reviewed, id_more], .5+0:5)
# --------------------------------------------------------------
```


You will now implement the cost function for collaborative filtering. Specifically, you should complete the following code to return J. 

```{r echo = TRUE}
cofiCostFunc <- function(Y, R, num_users, num_movies,
                         num_features, lambda = 0) {
  #COFICOSTFUNC Collaborative filtering cost function
  #   J <- COFICOSTFUNC(Y, R, num_users, num_movies, ...
  #   num_features, lambda)(params) returns the cost for the
  #   collaborative filtering problem.
  #
  # Notes: X - num_movies  x num_features matrix of movie features
  #        Theta - num_users  x num_features matrix of user features
  #        Y - num_movies x num_users matrix of user ratings of movies
  #        R - num_movies x num_users matrix, where R(i, j) <- 1 if the
  #            i-th movie was rated by the j-th user
    
  function(params) {
    # Unfold the U and W matrices from params
    X <-
      matrix(params[1:(num_movies * num_features)], num_movies, num_features)
    Theta <-
      matrix(params[(num_movies * num_features + 1):length(params)],num_users, num_features)
    
  
    # ----------------------- YOUR CODE HERE -----------------------
    J <- 0
    J <- (1 / 2) * sum(((X %*% t(Theta)) * R - Y * R) ^ 2) +
      (lambda / 2 * sum(Theta ^ 2)) + (lambda / 2 * sum(X ^ 2))
    # --------------------------------------------------------------
    J
  }
}
```

To help you debug your cost function, run the following code.

```{r echo = TRUE}
#  Load pre-trained weights (X, Theta, num.users, num.movies, num.features)
load("movieParams.Rda")
list2env(data,.GlobalEnv)
rm(data)

num_users <- as.numeric(num.users)
num_movies <- as.numeric(num.movies)
num_features <- as.numeric(num.features)

#  Reduce the data set size so that this runs faster
num_users <- 4; num_movies <- 5; num_features <- 3
X <- X[1:num_movies, 1:num_features]
Theta <- Theta[1:num_users, 1:num_features]
Y <- Y[1:num_movies, 1:num_users]
R <- R[1:num_movies, 1:num_users]

#  Evaluate cost function
J <- cofiCostFunc(Y, R, num_users, num_movies, num_features, 0)(c(c(X),c(Theta)))
           
cat(sprintf('Cost at loaded parameters: %f  (this value should be about 22.22)\n', J))
```

Once your cost function matches up with ours, you should now implement the collaborative filtering gradient function. Specifically, you should complete the following code to return the grad argument.

```{r echo = TRUE}
cofiGradFunc <- function(Y, R, num_users, num_movies, num_features, lambda = 0) {
  #cofiGradFunc returns the gradient for the
  #   collaborative filtering problem.
  # Notes: X - num_movies  x num_features matrix of movie features
  #        Theta - num_users  x num_features matrix of user features
  #        Y - num_movies x num_users matrix of user ratings of movies
  #        R - num_movies x num_users matrix, where R[i, j] <- 1 if the
  #            i-th movie was rated by the j-th user
  #
  function(params) {
    # Unfold the U and W matrices from params
    X <-
      matrix(params[1:(num_movies * num_features)], num_movies, num_features)
    Theta <-
      matrix(params[(num_movies * num_features + 1):length(params)],
             num_users, num_features)
    
    
    # You should set the following variables correctly:
    #        X_grad - num_movies x num_features matrix, containing the
    #                 partial derivatives w.r.t. to each element of X
    #        Theta_grad - num_users x num_features matrix, containing the
    #                     partial derivatives w.r.t. to each element of Theta
    #
    
    # ----------------------- YOUR CODE HERE -----------------------
    X_grad <- matrix(0,dim(X)[1],dim(X)[2])
    Theta_grad <- matrix(0, dim(Theta)[1], dim(Theta)[2])
    
    X_grad <- (((X %*% t(Theta)) * R) %*% Theta - (Y * R) %*% Theta) + lambda * X
    Theta_grad <- t((t(X) %*% ((X %*% t(Theta)) * R) - t(X) %*% (Y * R))) + lambda * Theta
    # --------------------------------------------------------------
    grad <- c(c(X_grad),c(Theta_grad))
    grad
  }
}
```

You can check your function by running the following code.

```{r echo = TRUE}
source("computeNumericalGradient.R")
source("checkCostFunction.R")
checkCostFunction()
```

(3 marks) Now, you should implement regularization for the cost function for collaborative filtering. Update the code of the *cofiCostFunc* function.

You can check your function using the following code.

```{r echo = TRUE}
lambda <- 1.5
J <- cofiCostFunc(Y, R, num_users, num_movies, num_features, lambda)(c(c(X),c(Theta)))
           
cat(sprintf('Cost at loaded parameters (lambda = 1.5): %f (this value should be about 31.34)\n', J))
```

(3 marks) Once your cost matches up with ours, you should proceed to implement regularization for the gradient. Update the code of the *cofiGradFunc* function.

You can check the gradient computations with the following code.

```{r echo = TRUE}
checkCostFunction(lambda)
```

Before training the collaborative filtering model, we will first add ratings that correspond to a new user that we just observed. This part of the code will allow you to put in your own ratings for the movies in our dataset!

```{r echo = TRUE}
#  Initialize my ratings
my_ratings <- rep(0,1682)

# Check the file movie_ids.txt for id of each movie in our dataset
# For example, Toy Story (1995) has ID 1, so to rate it "4", you can set
my_ratings[1] <- 4

# Or suppose did not enjoy Silence of the Lambs (1991), you can set
my_ratings[98] <- 2

# We have selected a few movies we liked / did not like and the ratings we
# gave are as follows:
my_ratings[7] <- 3
my_ratings[12]<- 5
my_ratings[54] <- 4
my_ratings[64]<- 5
my_ratings[66]<- 3
my_ratings[69] <- 5
my_ratings[183] <- 4
my_ratings[226] <- 5
my_ratings[355]<- 5

cat(sprintf('\n\nNew user ratings:\n'))

for (i in 1:length(my_ratings))
    if (my_ratings[i] > 0 )
        cat(sprintf('Rated %d for %s\n', my_ratings[i], movieList[i]))
```


Now we will train the collaborative filtering model.

```{r echo = TRUE}
load("movies_ratings.Rda")
list2env(data,.GlobalEnv) # "Y" and "R"
rm(data)

#  Add our own ratings to the data matrix

Y <- cbind(my_ratings, Y)
R <- cbind((my_ratings != 0), R)

#  Normalize Ratings
source("normalizeRatings.R")
NR  <- normalizeRatings(Y, R)
Ynorm <- NR$Ynorm
Ymean <- NR$Ymean
#  Useful Values
num_users <- dim(Y)[2]
num_movies <- dim(Y)[1]
num_features <- 10

# Set Initial Parameters (Theta, X)
n <- num_movies * num_features
X <- matrix(rnorm(n), num_movies, num_features)

n <- num_users * num_features
Theta <-  matrix(rnorm(n), num_users, num_features)

initial_parameters <- c(c(X), c(Theta))

# Set Regularization
lambda <- 10

cF <- cofiCostFunc(Ynorm, R, num_users, num_movies,num_features, lambda)
gF <- cofiGradFunc(Ynorm, R, num_users, num_movies,num_features, lambda)

#install.packages("lbfgsb3")
library(lbfgsb3)
source("lbfgsb3_.R")
theta <- lbfgsb3_(initial_parameters, fn= cF, gr=gF, control = list(trace=1,maxit=100))$prm

# The following code works but optim is slow on this problem
# theta <- optim(initial_parameters, fn = cF, gr = gF,
       #method = "BFGS", control = list(maxit=10, trace=1, REPORT=1) )$par

# Unfold the returned theta back into U and W
X <- matrix(theta[1:(num_movies*num_features)], num_movies, num_features)
Theta <- matrix(theta[(num_movies*num_features+1):length(theta)], num_users, num_features)

cat(sprintf('Recommender system learning completed.\n'))
```

(2 marks) After training the model, you have now computed X and Theta. Use them to compute the top 10 recommendations for the new user. Print the movie titles and the associated ratings.

```{r echo = TRUE}
# ----------------------- YOUR CODE HERE -----------------------
p <- X %*% t(Theta)
my_predictions <- p[,1] + Ymean

movieList <- loadMovieList()

ix <- sort(my_predictions, decreasing = TRUE,index.return=TRUE)$ix

cat(sprintf('\nTop recommendations for you:\n'))
for (i in 1:10){
    j <- ix[i]
    cat(sprintf('Predicting rating %.1f for movie %s\n', my_predictions[j],movieList[j]))
}

cat(sprintf('\n\nOriginal ratings provided:\n'))
for (i in 1:length(my_ratings))
    if (my_ratings[i] > 0 )
        cat(sprintf('Rated %d for %s\n', my_ratings[i],movieList[i]))
# --------------------------------------------------------------
```


# TURN IN 

- Your `.Rmd` file (which should knit without errors and without assuming any packages have been pre-loaded)
- Your Word (or pdf) file that results from knitting the Rmd.
- DUE: October 2, 11:55pm (late submissions not allowed), loaded into moodle.
